Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-15 manuscript/draft.Rmd
more Anna

These results raise a number of important issues. First, we should be skeptical
of predictions about future biodiversity from models that have not been
validated in terms of repeated observations at the same site. Since site-level
richness is relatively stable, model evaluations that show high accuracy across
spatial gradients may not indicate much about a model's ability to predict
changes over time. This issue has also recently been highlighted as an issue for
using SDMs to predict the future locations of individual species
[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
SDMs form the main foundation for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @urban2015].

Second, our results highlight the importance of comparing multiple modeling
approaches when conducting ecological forecasts, and in particular, the value of
comparing results to simple baselines to avoid over-interpreting the information
present in these forecasts (Box 1). Disciplines that have more mature
forecasting cultures often do this by reporting "forecast skill", i.e., the
improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
recommend following Ye et al. [@ye2015] and adopting this approach in future
ecological forecasting research. It also highlights the need for using
hindcasting and time-series data for assessing the effectiveness of forecasting
methods, since this allows comparisons to baselines and assessment of predicted
dynamics within sites or regions.

One of the main alternatives to stacked SDMs is modeling richness directly using
environmental variables and using the resulting models to make forecasts
[@algar2009; @distler2015]. Two comparisons of the the SDM and direct richness
modeling approaches reported that the methods yielded equivalent results for
forecasting diversity. While our results generally support rough equivalence
between stacked SDMs and direct richness modeling for point estimates, they also
show that stacked SDMs dramatically understimate the uncertainty. For this
reason, direct modeling of richness therefore is a better overall approach to
forecasting richness. A similar result is seen when comparing joint species
distribution models (JSDMs) to stacked single species distribution models. For
point estimates the joint distribution models are roughly equivalent to stacked
SDMs, but the JSDMs provide much better estimates of uncertainty [@harris2015;
others?]. In fact, JSDMs and direct richness modeling provide some of the best
estimates of uncertainty across all modeling approaches. This highlights the
importance of evaluating models based on their uncertainty as well as their mean
estimates (Box 1). Not doing so is particularly problematic in this case because
the summed species distribution models are overly confident meaning that using
them would lead to believing that richness would be restricted to a much
narrower range than would actually occur.

It is also possible that models that include species' relationships to their
environments or direct environmental constraints on richness may continue to
produce forecasts that are worse than simply assuming the systems are static.
This would be expected to occur if richness in these systems is not changing
over the relevant time scales and therefore simple models of no directional
change are most appropriate. Recent suggestions that local scale richness in
some systems is not changing directionally at multi-decadal scales supports this
possibility [@brown2001; @ernest2001; @dornelas2014; @vellend2013]. This lack of
change may be expected even in the presence of substantial changes in
environmental conditions and species composition at a site due to the
replacement of species from the regional pool [@brown2001; @ernest2001]. On
average the Breeding Bird Survey sites used in this study show little change in
richness (site-level SD of `r numbers$resid_sd` species, after controlling for
differences among observers) [see also @lasorte2005]. This absence of rapid
change is beneficial for the absolute accuracy of forecasts across sites,
because if a past year's richness is known it is easy to estimate a future
richness, and explains why stacked SDMs perform relatively well at this task
despite failing to capture meaningful dynamics. However it makes it difficult to
improve forecasts relative to simple baselines, since those baselines are close
to representing what is actually occurring in the system. This suggests that
simple time-series models and baselines should be actively considered for
forecasts of richness and other stable aspects of biodiversity and also suggests
that future efforts to understand and forecast biodiversity should also focus on
composition since it is expected to be more dynamic [@ernest2001;
@dornelas2014].

In addition to consideration of the different process models used for
forecasting it is important to consider the observation models. When working
with any ecological dataset there are imperfections in sampling that have the
potential to influence results. With large scale survey and citizen science
datasets like the Breeding Bird Survey these issues are potentially magnified by
the large number of different observers and major differences in habitat and
species [@sauer1994]. We included an observation model for one of the two major
observation issues known for the Breeding Bird Survey, differences among
observers. Accounting for differences in observers reduced the average error in
point estimates and also improved the coverage of the confidence intervals. In
addition, it resulted in changes in which models performed best, most notably
reducing the relative performance of the naive model for point estimates. This
suggests that the naive model performed well in part because it was capable of
accommodating rapid shifts in estimated richness introduced by changes in the
observer. These kinds of rapid changes were difficult for the other time-series
models to accommodate and so the average performance of the ARIMA and average
models improved once this source of observation error was addressed. This
demonstrates that properly modeling observation error can be important for
properly estimating and reducing uncertainty in forecasts and can also lead to
changes in the best methods for forecasting. We did not address differences in
detection probability across species and sites [@boulinier1998] since there is
no clear way to address this issue without making strong assumptions about the
data (i.e., assuming there is no biological variation in stops along a route),
but this would be a valuable addition to future forecasting models.

Future efforts will also need to begin to address the additional uncertainty
that comes from error in forecasting the environmental conditions themselves. In
this and other hindcasting studies the environmental conditions for the "future"
are known because the data has already been observed. In real forecasts the
environmental conditions themselves have to be forecast and those forecasts have
uncertainty and bias. This will cause ecological forecasts that use
environmental data to be more uncertain than observed in current hindcasting
efforts. It is important to correctly incorporate this uncertainty in the
predictor variables into forecasting models [@clark2001; @dietze2017]. Difficulty
in forecasting future environmental conditions at small scales will present
continued challenges for models incorporating these conditions and this may
result in a continued advantage to simple time-series based approaches.

Another challenge for ecological forecasting is the timescales of typical
forecasts. We evaluated forecasts using yearly timesteps up to a decade into the
future. In contrast, many ecological forecasts are temporally aggregated to 5-30
year timesteps and projected up to a century into the future. These are commonly
beyond the career or even lifespan of the researchers, making reasonable model
assessment impossible. Currently assessments at these long time-scales can
only be made with a small number of opportune datasets. For examples hindcasts
have been used to assess models of species richness [@algar2009, @distler2015]
and distribution [@rapacciuolo2012, @moran-ordonez2017, @araujo2005,
@eskildsen2013] using data aggregated across similar time scales. More
evaluation studies like these are needed but there is a paucity of adequate long
term data available. Short term forecasts will allow for relatively rapid
validation of models by researchers without needing to wait for enough data to
accumulate [@tredennick2016; @dietze2016]. This does not mean that they should
replace large scale long-term forecasts. As discussed above, drivers of species
richness are likely to differ at different temporal scales [@rosenzweig1995,
@white2004, @white2007]. Thus short-term forecasts should be expected to add
value to end users and inform, not replace, long-term forecasts. Hopefully by
evaluating trends in the shorter term forecasts as done here it will be
possible to also use the information from short-term forecasting to improve
longer-term estimates.

The science of forecasting biodiversity, and ecology more broadly, remains in
its infancy and it is important to consider the general inability of
forecasting methods to improve on simple baselines in that context. When weather
forecasting first started the forecasts were likewise worse than simple
baselines [@mcgill2012]. One of the things that helped weather forecasts improve
was large numbers of forecasts were made in public, which allowed different
approaches to forecasting to be regularly assessed and improved [@mcgill2012;
@silver2012]. This suggests that it is important for ecologists to start
regularly making and evaluating true ecological forecasts, even if they perform
poorly, and to make these forecasts publicly available for assessment. These
forecasts should include both short-term predictions, which can be assessed
quickly, and mid to long-term forecasts to help assess long time-scale processes
and determine how far into the future we can successfully forecast
[@tredennick2016; @dietze2017]. Forecasts based on the models in this paper from
now until 2050 are openly archived on Dryad so that we and others can assess how
well they perform and we plan to evaluate these forecasts and report the results
as each new year of BBS data becomes available. Weather forecasting has
continually improved throughout its history [@bauer2015] in part due to making
and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
do the same.

>>>>

These results raise a number of important issues. First, we should be skeptical
of predictions about future biodiversity from models that have not been
validated in terms of repeated observations at the same site. Since site-level
richness is relatively stable [[refs]], model evaluations that show high
accuracy across spatial gradients may not indicate much about a model's ability
to predict changes over time. This issue has also recently been highlighted as a
potential downside of using SDMs to predict the future locations of individual
species [@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering
because SDMs form the main foundation for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @urban2015].

Second, our results highlight the importance of comparing multiple modeling
approaches when conducting ecological forecasts, and in particular, the value of
comparing results to simple baselines to avoid over-interpreting the information
present in these forecasts (Box 1). Disciplines that have more mature
forecasting cultures often do this by reporting "forecast skill", i.e., the
improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
recommend following the advice of @ye2015 and adopting this approach in future
ecological forecasting research. Our results also highlight the need for using
hindcasting and time-series data for assessing the effectiveness of forecasting
methods. This practice allows comparisons to baseline models and also
facilitates the assessment of predicted dynamics within sites or regions.

One of the main alternatives to stacked SDMs is modeling richness directly using
environmental variables and using the resulting models to make forecasts. Two
studied compared the SDM and direct richness modeling approaches, and reported
that the methods yielded equivalent results for forecasting diversity
[@algar2009; @distler2015]. While our results generally support the idea of
rough equivalence between stacked SDMs and direct richness modeling for point
estimates, they also show that stacked SDMs dramatically understimate the
uncertainty. For this reason, direct modeling of richness is a better overall
approach to forecasting richness. A similar result is seen when comparing joint
species distribution models (JSDMs) to stacked single species distribution
models. For point estimates, the joint distribution models are roughly
equivalent to stacked SDMs, but the JSDMs provide much better estimates of
uncertainty [@harris2015; others?]. In fact, JSDMs and direct richness modeling
provide some of the best estimates of uncertainty across all modeling
approaches. This highlights the importance of evaluating models based on their
uncertainty as well as their mean estimates (Box 1). Not doing so is
particularly problematic in this case, because the summed species distribution
models are overly confident, which means that these models could mislead
ecologists into believing that richness would be restricted to a much narrower
range than would actually occur.

It is also possible that models that include species' relationships to their
environments or direct environmental constraints on richness will continue to
produce forecasts that are worse than simply assuming the systems are static.
This would be expected to occur if richness in these systems is not changing
over the relevant time scales (e.g. 10-30 years in this study), which would make
simpler models with no directional change more appropriate. Recent suggestions
that local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@dornelas2014; @vellend2013]. This lack of change in richness may be expected
even in the presence of substantial changes in environmental conditions and
species composition at a site due to the replacement of species from the
regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
sites used in this study show little change in richness [site-level SD of `r
numbers$resid_sd` species, after controlling for differences among observers;
see also @lasorte2005]. The absence of rapid change in this dataset is
beneficial for the absolute accuracy of forecasts across different sites: when a
past year's richness is already known, it is easy to estimate a future richness.
The site-level stability of the BBS data also explains why stacked SDMs perform
relatively well at predicting future richness, despite failing to capture
changes in richness over time. However, this stability also makes it difficult
to improve forecasts relative to simple baselines, since those baselines are
already close to representing what is actually occurring in the system. These
results suggest that simple time-series models and baselines should be actively
considered for forecasts of richness and other stable aspects of biodiversity.
Our results also suggest that future efforts to understand and forecast
biodiversity should also focus on species composition, since lower-level
processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
contain more useful information [@harris2015].

**Observer effects.** In addition to consideration of the different process
models used for forecasting it is important to consider the observation models.
When working with any ecological dataset, there are imperfections in the
sampling process that have the potential to influence results. With large scale
surveys and citizen science datasets, such as the Breeding Bird Survey, these
issues are potentially magnified by the large number of different observers and
by major differences in the habitats and species being surveyed [@sauer1994].
Accounting for differences in observers reduced the average error in our point
estimates and also improved the coverage of the confidence intervals. In
addition, controlling for observer effects resulted in changes in which models
performed best, most notably reducing the relative performance of the naive
model's point estimates. This suggests that, prior to accounting for observer
effects, the naive model performed well because it was capable of accommodating
rapid shifts in estimated richness introduced by changes in the observer. These
kinds of rapid changes were difficult for the other time-series models to
accommodate, and so the average performance of the ARIMA and average models
improved once this source of observation error was addressed. This demonstrates
that modeling observation error can be important for properly estimating and
reducing uncertainty in forecasts and can also lead to changes in the best
methods for forecasting. We did not address differences in detection probability
across species and sites [@boulinier1998] since there is no clear way to address
this issue without making strong assumptions about the data (i.e., assuming
there is no biological variation in stops along a route), but this would be a
valuable addition to future forecasting models.

**Environmental uncertainty.** Future biodiversity forecasting efforts will also
need to begin to address the additional uncertainty introduced by the error
inherent in the process of forecasting the environmental conditions that
ecologists use as predictor variables. In this, and other hindcasting studies,
the environmental conditions for the "future" are known because the data has
already been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts.
Thus, it is important to correctly incorporate this uncertainty in the predictor
variables into forecasting models [@clark2001; @dietze2017], as we did with the
uncertainty in observer-level effects. Difficulty in forecasting future
environmental conditions at small scales will present continued challenges for
models incorporating environmental variables, and this may result in a continued
advantage for simple time-series based approaches.

**Multi-timescale forecasts.** Another challenge for ecological forecasting is
the timescales of typical forecasts. We evaluated forecasts using yearly
timesteps up to a decade into the future. In contrast, many ecological forecasts
are temporally aggregated to 5-30 year timesteps and projected up to a century
into the future. These are commonly beyond the career or even lifespan of the
researchers, which makes it impossible to directly validate the models'
predictions. Currently, assessments at these long time-scales can only be made
with a small number of opportune datasets. For example, hindcasts have been used
to assess models of species richness [@algar2009; @distler2015] and distribution
[@rapacciuolo2012; @moran-ordonez2017; @araujo2005; @eskildsen2013] using data
aggregated across similar time scales to those studied here. More evaluation
studies like these are needed but there is a paucity of adequate long term data
available. Short term forecasts will allow for relatively rapid validation of
models by researchers without needing to wait for enough data to accumulate
[@tredennick2016; @dietze2016]. This does not mean that they should replace
large scale long-term forecasts. As discussed above, drivers of species richness
are likely to differ at different temporal scales [@rosenzweig1995; @white2004;
@white2007]. Thus, short-term forecasts should be expected to add value to
scientists and policymakers, and inform (not replace) long-term forecasts.
Hopefully, by evaluating trends in the shorter term forecasts as done here it
will be possible to also use the information from short-term forecasting to
improve longer-term estimates.

**Conclusions.** The science of forecasting biodiversity, and ecology more
broadly, remains in its infancy and it is important to consider the general
inability of forecasting methods to improve on simple baselines in that context.
When weather forecasting first started the forecasts were likewise worse than
simple baselines [@mcgill2012]. One of the things that helped weather forecasts
improve was large numbers of forecasts were made in public, which allowed
different approaches to forecasting to be regularly assessed and improved
[@mcgill2012; @silver2012]. This suggests that it is important for ecologists to
start regularly making and evaluating true ecological forecasts, even if they
perform poorly, and to make these forecasts publicly available for assessment.
These forecasts should include both short-term predictions, which can be
assessed quickly, and mid to long-term forecasts to help assess long time-scale
processes and determine how far into the future we can successfully forecast
[@tredennick2016; @dietze2017]. Forecasts based on the models in this paper from
now until 2050 are openly archived on Dryad so that we and others can assess how
well they perform and we plan to evaluate these forecasts and report the results
as each new year of BBS data becomes available. Weather forecasting has
continually improved throughout its history [@bauer2015] in part due to making
and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
do the same.
